---
title: Learning data.table (3)
author: Rui Ying
date: '2021-04-29'
slug: []
categories:
  - R
tags: []
subtitle: ''
summary: 'Day 3: use key in data.table'
authors: []
lastmod: '2021-04-29T21:17:05+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
# What's key

Key is very similar to the group_by concept, or the telephone number book in our world: when you want to find someone's phone number, you go find his/her first name, then second name. The names here are key.

Or as the data.table vignette shows, the `key` in data.table is inherited from `rowname` in data.frame. However, `key` is advantageous in

- each row can have many keys in dt, but only one rowname in df
- keys are not unique (you can have duplicate variable in  key column)
- keys can be in various variable types
- key columns are well sorted

# How to set key
setkey() or setkeyv(): the former one is better in interactive use, whilst the latter one is more of function-use.

```{r eval=FALSE}
setkey(flights, origin, dest)
#is equal to
setkeyv(flights, c("origin", "dest"))
```

# Feature of using key

- it modify data.table and return result invisibly
- it's manipulating by reference as `:=` operator and all `set*` family function (setkey, setname etc.)

# Do some work 
```{r eval=FALSE}
setkey(flights, origin, dest)
# select j
flights[.("LGA", "TPA"), .(arr_delay)]

# do sth in j
flights[.("LGA", "TPA"), max(arr_delay)]

# use by
flights["JFK", max(dep_delay), keyby = month]

```
# `mult` and `nomatch` argument

Very simply, `mult` choose how many matching rows to return, and `nomatch` selects if the unmatching NA should be returned or skipped.

The default value of these two are "all" and "NA". Setting `nomatch=NULL` kips queries with no matches.

# Ad of using key
Key is based on binary search (二分法) with O(Log(N)) complexity, while the traditional vector scan method is simply scaning as the name shows. The complexity is O(N).

But the document says the vector scan has been optimized in recent version thus is also binary search-based, otherwise you set the key to `NULL` and then it's back to the slow one.

