---
title: Learning data.table (4)
author: Rui Ying
date: '2021-04-30'
slug: []
categories:
  - R
tags: []
subtitle: ''
summary: 'Day 4: reshape data.table'
authors: []
lastmod: '2021-04-30T16:38:36+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

# `melt` and `dcast`

Actually I don't know the difference of these two functions between data.table and reshape2, but the use should be same.

## `melt` converts wide to long

```{r}
library(data.table)
s1 <- "family_id age_mother dob_child1 dob_child2 dob_child3
1         30 1998-11-26 2000-01-29         NA
2         27 1996-06-22         NA         NA
3         26 2002-07-11 2004-04-05 2007-09-02
4         32 2004-10-10 2009-08-27 2012-07-21
5         29 2000-12-05 2005-02-28         NA"
DT <- fread(s1)
DT
str(DT)

DT.melt <- melt(DT, id.vars = c("family_id","age_mother"),
                measure.vars = c("dob_child1", "dob_child2", "dob_child3"),
                variable.name = "child",
                value.name = "dob")
head(DT.melt)
```

- If neither `id.var` and `measure.var` are specified , the `id` will assign all non-integer, logical, numeric columns.
- By default, var and value name is simply `var` and `value`.

## `dcast` converts long to wide

dcast use formula as below
```{r}
dcast(DT.melt, family_id + age_mother ~ child, value.var = "dob") #value.var is the variable to fill all cells
```
# Limitation in combine multiple columns
```{r}
s2 <- "family_id age_mother dob_child1 dob_child2 dob_child3 gender_child1 gender_child2 gender_child3
1         30 1998-11-26 2000-01-29         NA             1             2            NA
2         27 1996-06-22         NA         NA             2            NA            NA
3         26 2002-07-11 2004-04-05 2007-09-02             2             2             1
4         32 2004-10-10 2009-08-27 2012-07-21             1             1             1
5         29 2000-12-05 2005-02-28         NA             2             1            NA"
DT <- fread(s2)
DT
```


## An untidy way to implement this (not run)
```{r eval=FALSE}
DT.m1 = melt(DT, id = c("family_id", "age_mother"))
DT.m1[, c("variable", "child") := tstrsplit(variable, "_", fixed = TRUE)] #I don't even understand this weird function tstrsplit
DT.c1 = dcast(DT.m1, family_id + age_mother + child ~ variable, value.var = "value")
DT.c1
```

## Enhanced ways

### Pass a list to measure.vars where each element of the list contains the column that should be combined together.
```{r}
colA = paste("dob_child", 1:3, sep = "")
colB = paste("gender_child", 1:3, sep = "")
DT.m2 = melt(DT, measure = list(colA, colB), value.name = c("dob", "gender")) #notice the measure paramter instead of measure.var
DT.m2
```

The additional `variable` column is good to do dcast later and *can be removed* if necessary!

### Using `patterns()`
```{r}
DT.m2 = melt(DT, measure = patterns("^dob", "^gender"), value.name = c("dob", "gender"))
DT.m2
#the ^ here means letters beginning with "dob" and "gender"
#similarly, _ means end with xxx
```

You can also do this using `stat::reshape()` in base R

### Enhanced `dcast`

```{r}
DT.c2 <- dcast(DT.m2, family_id + age_mother ~ variable, value.var = c("dob", "gender") )
DT.c2
```
### 
