---
title: Learning data.table (2)
author: Rui Ying
date: '2021-04-25'
slug: []
categories:
  - R
tags: []
subtitle: ''
summary: 'Day 2: data.table add/update/delete columns by reference'
authors: []
lastmod: '2021-04-25T22:35:28+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---
```{r}
#download data as usual
library(data.table)
library(magrittr)
file_url <- 'https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv'
df <- fread(file_url)
```
# Introduction
This post studies the use of `:=` operator which add/update/delete *columns*. The difference between `:=` and default data.frame operation is the former does by reference and therefore is not deep copy and not assign any value to a variable. This causes faster speed, less memory use but some side effects.

## Two forms of `:=` ways
```{r eval=FALSE} 
# LHS := RHS
DT[, c("colA", "colB") := list(valA, valB)]

# function form
DT[, `:=`(colA = valA, # valA is assigned to colA
          colB = valB)]
```

## Add/update/delete columns
```{r}
df[, `:=`(speed = distance / (air_time/60),
   delay = arr_delay + dep_delay)] %>% head()
```
# Subassign - conditional update
```{r}
df[hour == 24L, hour := 0L][] #replace all hour 24 to 0
# note := returns the result invisibly, so we have extra [] here
```

# Delete columns
```{r}
df[, c("delay") := NULL]
```

# Using `by`
```{r}
df[, max_speed := max(speed), by = .(origin, dest)]
```

# Multiple columns
```{r}
in_cols  = c("dep_delay", "arr_delay")
out_cols = c("max_dep_delay", "max_arr_delay")
df[, c(out_cols) := lapply(.SD, max), by = month, .SD = in_cols][] %>% head()

#note the c() in out_cols because we can not do out_cols := lapply(.SD, max). That would result in adding one new column named out_col.
```

# Side effects of `:=`
Because `:=` refer to the memory directly, it can be dangerous in writing function: any operation would change the origin data set. Thus a better way is use `copy()` first and do other things in the duplicate one.
