---
title: Learning data.table (1)
author: Rui Ying
date: '2021-04-23'
slug: []
categories:
  - R
tags: []
subtitle: ''
summary: 'Day 1: data.table syntax'
authors: []
lastmod: '2021-04-23T23:13:43+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
runtime: shiny
---
# Prepare data
```{r}
library(data.table)
file_url <- 'https://raw.githubusercontent.com/Rdatatable/data.table/master/vignettes/flights14.csv'
df <- fread(file_url)
head(df,3)
```
# Simple syntax

data.table use one-line syntax like `df[i, j, by]` where `i` is the rows to select and to manipulate, `j` is the columns to select and how to manupulate it (e.g., mutate function in dplyr), `by` is the group option.

# Subset rows
```{r}
library(magrittr)
#using condition
sub_df <-  df[year==2014 & hour < 10,]
head(sub_df)
#using index (not recommended because of bad readability and maintainance)
sub_df <-  df[1:5,]
head(sub_df)
#i parameter can sort specific columns
df[order(hour,-distance)] %>% head() #`-` means descend order
```
# Subset columns
```{r}
#using list
df[, .(dep_delay, arr_delay)]

#which is same as
df[, list(dep_delay, arr_delay)]

# or 
df[, c("dep_delay", "arr_delay")]

# or 
list <- c("dep_delay", "arr_delay")
df[, ..list] #use .. here from the convention of unix directory operation
#but I understand in a better way: . means list, .. means defining and call this (two steps)

# parameter *with*
df[, list, with=FALSE]
#with() is a functional programming style so we can treat columns as variable. with=FALSE means restoring to data.frame manipulation. with=TRUE is default

#if there's only 1 col, the .() can be omitted
df[, arr_delay] %>% head() #a vector as the column itself

#we can also use index here
df[, year:day] #var1 to var2

#deselect using - or !
df[, !c("arr_delay", "dep_delay")]
#or df[, -c("arr_delay", "dep_delay")]

#deselect using index
df[, !(year:day)]
```

# Do something with `j`
```{r}
#rename
df[, .(delay_arr = arr_delay, delay_dep = dep_delay)] %>% head()

#numbers .N
df[origin == "JFK" & month == 6L, .N] %>% head()
#.N is a special built-in variable that holds the number of observations in the current group.

#a same way
df[origin == "JFK" & month == 6L, length(dest)]

#calculate mean value
df[origin == "JFK" & month == 6L,
   .(m_arr = mean(arr_delay), m_dep = mean(dep_delay))] %>% head()
```
# Aggregation using `by`
```{r}
# by is similar to group_by() in tidyverse
# I imagine it as cut a large block into many small ones
# by will be the key index in output, which should be quite familiar with pandas user

# a simple example
df[, .N, by = origin] %>% head()

# multiple group
df[carrier == "AA", .N, by = .(origin, dest)] %>% head()

# calculate in j
df[carrier == "AA",
   .(mean(arr_delay), mean(dep_delay)),
   by = .(origin, dest, month)] %>% head()

#sorted_by: keyby
#nothing different but in a tidy way
df[carrier == "AA",
   .(mean(arr_delay), mean(dep_delay)),
   keyby = .(origin, dest, month)] %>% head()

#expression in by
df[, .N, .(dep_delay>0, arr_delay>0)] %>% head()

```



# What's `.SD`?

`.SD` means *S*elect of *D*ata. It's a data.table that holds the data for the current group defined using `by`. I prefer regrading it as "grouped data".
  
```{r}
#.SD
DT = data.table(
  ID = c("b","b","b","a","a","c"),
  a = 1:6,
  b = 7:12,
  c = 13:18)
print(DT)
DT[, print(.SD), by = ID]

#calculate each block's mean
DT[, lapply(.SD, mean), by = ID]
```
So you can find `.SD` contains all columns as origin `DT` but without the grouping column `ID`. And `.SD` is well sorted.

# `.SDcol` specify what to compute
```{r}
df[carrier == "AA",                       ## Only on trips with carrier "AA"
   lapply(.SD, mean),                     ## compute the mean
   by = .(origin, dest, month),           ## for every 'origin,dest,month'
   .SDcols = c("arr_delay", "dep_delay")] ## for just those specified in .SDcols
```

# subset `.SDcol`
```{r}
# return first 2 cols for each month
# you can always treat `by` as "for each"
df[, head(.SD, 2), by=month] %>% head()
```
