---
title: Review data analysis in R
author: Rui Ying
date: '2021-01-06'
slug: []
categories:
  - R
tags:
  - R; Data Science
subtitle: ''
summary: 'tidyverse, reshap2 functions review'
authors: [Rui Ying]
lastmod: '2021-01-06T17:12:47Z'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

# Use tibble to replace "old" data.frame

## Differences:
- print() by specifying rows; 
- subset (you can do slice like in Pandas).
```{r warning=FALSE,message=FALSE}
library(tidyverse)
library(magrittr)
#mtcars %>% print(n=10,width = Inf) #always print all columns, regardless of the width of the screen.
mtcars[[1]] == mtcars[['mpg']] #first column/variable
mtcars[,1] == mtcars$mpg #the same
mtcars %>% select(mpg) %>% head(5) #select() is also used in subset but a little different
```
## convert with data.frame
```{r}
df <- tibble(foo = 1, bar = 'bar')
class(as.data.frame(df)) #[1] "data.frame"
class(as_tibble(df)) #[1] "tbl_df"     "tbl"        "data.frame"
```
# Reshape data

## What is Tidy data
Column -> variable, row -> observation.
Ad:
1. uniformity; 2. Placing variable in columns (or a "vector" in R) allows R and tidyverse to work naturally.
e.g.
```{r}
mutate(mtcars,new_col = cyl+1) %>% head(5)
```

## dplyr basics
- Pick *observations* by their values: filter()
- Reorder the rows: arrange()
- Pick *variables* by their names: select()
- Create new variable by computing based on existing variables: mutate()
- A simple summary: summarise()
- Group data by one or more variables: group_by() (always ungroup() after grouping)

## Pivot
### Longer table
Question: *Column names are not names of variables, but values of a variable. For example:*
```{r}
head(table4a)
```
We can divide the two untidy columns into a variable column (with col_name 'year') and a value column (with col_name 'cases')
```{r}
table4a %>% pivot_longer(c('1999','2000'), names_to = "year", values_to = "cases")
table4a %>% gather(`1999`,`2000`,key = 'year',value = 'cases') #the gather() in tidyr is equal
```
### Wider pivot
Question: *Observations is scattered across multiple rows* e.g. the cases and population are both observation, which should be put in one row.
```{r}
 head(table2,5)
```
```{r}
table2 %>% pivot_wider(names_from = type, values_from = count) # here's 2 paramters only
table2 %>% spread(type,count) #spread() is equal as well
```
## Joins
The *variables* used to connect each pair of tables are called *keys*, if it is *primary key* in one table (i.e. define the observation or row name), then it's a foreign key in other tables. *Primary key* are not necessary for a table.

Four *mutate* joins (__add__ new matched variable to primary): inner (smallest subset), left (as primary), right (as primary), full (collection).
Two *filter* joins (__filter__ matched observations to save or drop).

## reshap2 package

### melt()
`melt()` breaks data into a data.frame with only two columns (variable and value), which is exactly transforming a "wide format" data into a "long format" data.
parameters:
- `id`
- `variable.name` (change the default variable column name)
- `value.name` (change the default value column name)
```{r warning=FALSE,message=FALSE}
library(reshape2)
table1_melt <- table1 %>% melt(id=c("country","year"))
table1_melt
```
### dcast()
`dcast()` do the opposite thing as `melt()`, transforming long to *wide data*. Additional parameter including `mean`,
The difference between `dcast` and `acast` is the output in data.frame (2-D only) or vector/array/matrix.
```{r}
table1_melt %>% dcast(country+year~variable) #turn it back
```


This post is built by rmarkdown. Seemingly Rstudio and Rmd are good tool to manage this site.
